---
title: "Electric Project"
author: "Caleb Waack"
date: "2023-01-11"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Electric Project 

# Overview

The purpose of this analysis is build skills in R. Personally I have lots of experience in Python, and some experience in R. Although I have completed many projects in R via my masters program. I have not done any projects for fun. Which would indicate that I haven't developed the skills needed to work with messy data in R. 

The goal of this analysis is:

1.) What is the primary power source for each country? 
2.) How has the primary power source changed over time? 
3.) Who are the top 5 consumers of each energy source? 
4.) By percent kW/hr how much renewable energy has been used by each country over the past 10 years? 
5.) Are there any common threads in the data that indicate a transition point between renewable energy resources and fossil fuels? 
6.) Who are the top 5 energy consumers in this data set? What are their sources? How have they changed in the past 10, 20, 30 years? 
7.) Predicted the energy consumption for the top 10 energy consumers for the next 10 years? How do the energy sources change? 

```{r,error=FALSE}
#First create a block of code to install all of the packages
library(readr)
library(dplyr)
```

# Task 1:

Load and familiarize yourself with the data. 

We are looking to answer questions like:
How many missing values are there for each column? How many for each country? Should we consider dropping some countries or years? 
What is the distribution of non-missing data? Is it normal? How much does it vary year to year for the same country? Or even countries in close proximity to each other? 

```{r}
#First we load the data
setwd('C:/Users/caleb/OneDrive/Desktop/Personal/Data Science/Personal Projects/Pair Programming/')
data <- read_csv('World Energy Consumption.csv',show_col_types = FALSE)
#Next calculate summary stats for each column and country. 
```

```{r}
summary(data$country)
length(unique(data$country))
#this answers the number of na values in a column
sum(is.na(data$coal_prod_change_pct))
```
```{r}
v <- colnames(data)
```


```{r}
for (ii in colnames(data)) {
  print(str(ii))
  value <- sum(is.na(data[[ii]]))
  print(str(value))
}
```

# Task 2

We want to know the different components of energy consumption. Separate the data into two different dataframes, one for renewable energies, and the other for non-renewables. 
```{r}

```






