---
title: "Electric Project"
author: "Caleb Waack"
date: "2023-01-11"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Electric Project 

# Overview

The purpose of this analysis is build skills in R. Personally I have lots of experience in Python, and some experience in R. Although I have completed many projects in R via my masters program. I have not done any projects for fun. Which would indicate that I haven't developed the skills needed to work with messy data in R. 

The goal of this analysis is:

1.) What is the primary power source for each country? 
2.) How has the primary power source changed over time?
3.) Who are the top 5 consumers of each energy source? 
4.) By percent kW/hr how much renewable energy has been used by each country over the past 10 years? 
5.) Are there any common threads in the data that indicate a transition point between renewable energy resources and fossil fuels? 
6.) Who are the top 5 energy consumers in this data set? What are their sources? How have they changed in the past 10, 20, 30 years? 
7.) Predicted the energy consumption for the top 10 energy consumers for the next 10 years? How do the energy sources change? 

```{r,error=FALSE,warning=FALSE}
#First create a block of code to install all of the packages
library(readr)
library(dplyr)
library(imputeTS)
library(ggplot2)
library(factoextra)
library(tidyverse)
```

## Task 1:
Load and familiarize yourself with the data. 

We are looking to answer questions like:
How many missing values are there for each column? How many for each country? Should we consider dropping some countries or years? 
What is the distribution of non-missing data? Is it normal? How much does it vary year to year for the same country? Or even countries in close proximity to each other? 

```{r}
#First we load the data
setwd('~/GitHub/Electric-Project/')
data <- read_csv('World Energy Consumption.csv',show_col_types = FALSE)
#Next calculate summary stats for each column and country. 
```

```{r}
summary(data$country)
length(unique(data$country))
#this answers the number of na values in a column
sum(is.na(data$coal_prod_change_pct))
```

```{r,echo=FALSE}
# for (ii in colnames(data)) {
#   print(str(ii))
#   value <- sum(is.na(data[[ii]]))
#   print(str(value))
# }

#remove all na values from the data
# data <- data %>%
#   na.exclude()

# for (ii in colnames(data)){
#   data <- data %>%
#     filter(!is.na(ii))
# }

# data <- na_replace(data,0)
```
Filter the data by the large groupings, keep in mind there are "other countries"

```{r,echo=FALSE}
unique(data$country)
#Asia Pacific, Africa, North America, Europe, South & Central America
```
Create continents list

```{r,echo=FALSE}
continents <- list('Asia Pacific'
               , 'Africa'
               , 'North America'
               , 'Europe'
               ,'South & Central America')
```
Filter on continents list
Filter on year

```{r,echo=FALSE}
# filter function works like
# filter(dataset_name,column1==value,column2 >= value)
# you can put as filter on as many columns as you want


# continents_data <- filter(data, country %in% continents)
# continents_data <- filter(continents_data, year >= 1950)

continents_data <- data %>%
  filter(country %in% continents) %>%
  filter(year >=1950)

head(continents_data,10)
```


## Basic visuals:
-Histogram is something that I use in almost every analysis
-It tells us how widely the data is distributed, and if the data is skewed or not
```{r}

ggplot(continents_data, aes(x=coal_prod_change_twh)) + geom_histogram()

```
```{r}
count(continents_data,coal_prod_change_twh,sort=TRUE)
```
```{r}
# continents_data <- filter(continents_data,country=='North America')
ggplot(continents_data, aes(x=year,
                            y=coal_prod_change_pct,
                            group=country))+ geom_line(aes(color=country))
```



## Task 2:
We want to know the different components of energy consumption. Separate the data into two different dataframes, one for renewable energies, and the other for non-renewables. 
```{r}
# select dplyr function
# select(dataset_name, column1, column2,column_n)
columns <- colnames(continents_data)

renewables <- select(continents_data,
                     country,
                     year,
                     columns[101:107],
                     columns[91:97],
                     population)

nonrenewables <- select(continents_data,
                        country,
                        year,
                        columns[4:9],
                        columns[20:28],
                        columns[44:59],
                        columns[82:90],
                        population)
```

## Task 3:
Who are the top 5 consumers of renewable non-renewable energy source?
```{r}
# arrange dplyr function
# arrange(dataset_name, col1,col2, col_n)
# default order is asc, to change to desc use the desc function, i.e. desc(col_n)
head(arrange(renewables,desc(renewables_share_energy)))

renewables_sorted <- renewables %>%
  select(country,year,renewables_share_energy) %>%
  group_by(country) %>%
  mutate(avg_renewables = mean(renewables_share_energy,na.rm = TRUE)) %>%
  select(country,avg_renewables) %>%
  summarise(avg_renewables = mean(avg_renewables,na.rm = TRUE,.groups=country)) %>%
  arrange(desc(avg_renewables))
  
head(renewables_sorted)
```


```{r}
summary(filter(renewables,country=='Europe'))

ggplot(data = filter(renewables,country=='Europe'),
       aes(x=year,y=renewables_share_energy)) + geom_col(fill='blue',color='black') + labs(title='Europe Energy Renewable Energy Share')

ggplot(data = filter(renewables,country=='Africa'),
       aes(x=year,y=renewables_share_energy)) + geom_col(fill='red',color='black') + labs(title='Africa Energy Renewable Energy Share')

ggplot(data=filter(renewables,country=='North America'),
       aes(x=year,y=renewables_share_energy)) + geom_col(fill='purple',color='black') +
  labs(title='North America Renewable Energy Share')

ggplot(data=filter(renewables,country=='South & Central America'),
       aes(x=year,y=renewables_share_energy)) + geom_col(fill='orange',color='black') +
  labs(title='South & Central America Renewable Energy Share')

ggplot(data=filter(renewables,country=='Asia Pacific'),
       aes(x=year,y=renewables_share_energy)) + geom_col(fill='green',color='black') +
  labs(title='Asia Pacific Renewable Energy Share')



```

## Task 4:
By percent kW/hr how much renewable energy has been used by each country over the past 10 years?
```{r}
# mutate dplyr function
# mutate(dataset_name, new_col = col1 - col2)
renewables <- renewables %>%
  filter(year>=2009) %>%
  mutate(total_renewables = renewables_share_elec + other_renewables_share_elec)

renewables_sorted <- renewables %>%
  select(country,year,renewables_share_energy) %>%
  group_by(country) %>%
  mutate(avg_renewables = mean(renewables_share_energy,na.rm = TRUE)) %>%
  select(country,avg_renewables) %>%
  summarise(avg_renewables = mean(avg_renewables,na.rm = TRUE,.groups=country)) %>%
  arrange(desc(avg_renewables))

head(renewables_sorted)

```

## Task 5:
Are there any common threads in the data that indicate a transition point between renewable energy resources and fossil fuels? At what population, energy per capita, or energy per gdp do we see a transition to renewables? Descriptive
```{r}

```
## Task 6:
Who are the top 5 energy consumers in this data set? What are their sources? How have they changed in the past 10, 20, 30 years? Descriptive
```{r}
# summarise dplyr function
# summarise(dataset_name, col=function(column))
# summarise(dataset_name, new_col=sum(col1))
```

## Task 7:
Predict the energy consumption for the top 5 energy consumers for the next 10 years? How do the energy sources change? Predict only the top energy sources in the data. Predictive Time Series
```{r}

```

## Task 8:
Which countries have similar energy consumption habits over the past 20 years? Use unsupervised learning to segment the countries into logical groups. How many groups are there? Predicitve / Descriptive Segmentation
```{r}
# unique(continents_data$country)
#
# unique(data$country)

 data <- data %>%
   filter(year >= (max(data$year)-20))

# Commented out the code above because it was already filtered. 

countries_list <- list('United States',
                       'Austria',
                       'Canada',
                       'Norway',
                       'France',
                       'Germany',
                       'Portugal',
                       'Hong Kong',
                       'China',
                       'Australia',
                       'New Zealand',
                       'Thailand',
                       'Japan',
                       'Cameroon',
                       'Egypt',
                       'Djibouti',
                       'Honduras',
                       'Brazil')

countries_data <- data %>%
  filter(country %in% countries_list)

# Quick validation that we have the right number of unique values
length(unique(countries_data$country))

```
```{r}
summary(countries_data)

# filter through dataset by number of na values greater than 30% of the data
```

```{r}
filter_cols <- list('country','year')

for (v in colnames(countries_data)) {
  if (endsWith(v,'per_capita') == TRUE) {
    filter_cols <- append(filter_cols,v)
  }
}

filter_cols <- unlist(filter_cols)
```


```{r}
countries_data <- countries_data %>%
  select(all_of(filter_cols))
```

```{r}
scaled_data <- as.data.frame(scale(countries_data%>%select(-country,-year)))
```


```{r}
scaled_data$country <- countries_data$country
scaled_data[is.na(scaled_data)] <- 0
```

```{r}
cluster <- kmeans(scaled_data%>%select(-country), 3)

kmeans(scaled_data%>%select(-country),3)
```

```{r}
countries_data <- countries_data %>%
  mutate(class = cluster$cluster)
```


```{r}

ggplot(aes(x=renewables_energy_per_capita, y=energy_per_capita,col=class), data=countries_data) + geom_point()

```


```{r}
cluster$centers
```

```{r}
scaled_data <- scaled_data%>%select(-country)
```


```{r}
output <- map_dbl(1:10, function(k){
  model <- kmeans(x = scaled_data, centers = k)
  model$tot.withinss
})

# tot.withinss stands for the total within-cluster sum of squares

elbow_method <- data.frame(k = 1:10,
                           centers_values = output)

ggplot(elbow_method, aes(x = k, y = centers_values)) + 
  geom_line() + geom_point() + scale_x_continuous(breaks = 1:10)
```

```{r}

cluster <- kmeans(x=scaled_data, centers = 4)

countries_data <- countries_data %>%
  mutate(class = cluster$cluster)

fviz_cluster(cluster, data = scaled_data)
```

```{r}
# cluster <- kmeans(x=scaled_data, centers = 6)
# 
# fviz_cluster(cluster, data = scaled_data)
```

```{r}
cluster_centers <- cluster$centers
```

## Explanation:
A couple things to keep in mind here. 

All components of the kmeans model can be called off the model itself with a dollar sign, and each component is explained below.

Cluster: Is the class that the algorithm has assigned to the specific entry of the data.

Centers: Is the center of that classes cluster for that attribute in n-dimensional space.

Totss: Stands for the Total Sum of Squares. In short this is the sum of the differences of each sample from the mean of the population, squared. 

Withinss: Stands for the Within Sum of Squares. This tells us how much variation is in each cluster. If you have 4 clusters there should be 4 values. A smaller value indicates a more tightly grouped cluster. 

Tot.withinss: Is the total sum of squares within sum of squares. This is how we determine the ideal number of groups for our algorithm, and is the sum of the 4 values for withinss. The smaller the variation amongst the individual clusters the more likely they are similiar to each other. 

Betweenss: totss - tot.withinss, this tells us the distance between the clusters. Can be useful if we have clusters that are close together. 

Size: The number of points in each cluster. 


## How to leverage this analysis
Now that we have this created, what can we do with it?
1) What are the characteristics of each class? 

```{r}

agg_class <- countries_data %>%
  group_by(class) %>%
  summarise(energy_avg = mean(energy_per_capita,na.rm = TRUE),
            renewable_avg = mean(renewables_energy_per_capita, na.rm = TRUE),
            non_renewable_avg = mean(fossil_energy_per_capita, na.rm = TRUE))
  
ggplot(data = agg_class,aes(x=class,y=energy_avg)) +
         geom_col(fill='blue',color='black') +
  labs(title = 'Energy Consumption by Class')

ggplot(data = agg_class, aes(x=class,renewable_avg)) +
  geom_col(fill = 'green', color='black') + 
  labs(title = 'Renewable Energy Consumption by Class')

ggplot(data = agg_class, aes(x=class, non_renewable_avg)) +
  geom_col(fill='red',color='black') +
  labs(title = 'Non-Renewable Energy Consumption by Class')


```


```{r}
class_3 <- filter(countries_data,class==3)

class_3 %>%
  group_by(country) %>%
  count()

countries_data %>%
  group_by(country) %>%
  count()
```


```{r}
cor(countries_data$coal_prod_per_capita, countries_data$energy_per_capita,use = 'complete.obs', method='pearson')
```






